% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\documentclass[
]{article}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\usepackage{fullpage}
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\author{}
\date{}

\begin{document}

\section{Variational Autoencoders: Pre-reading
Notes}\label{variational-autoencoders-pre-reading-notes}

We've spent the year building neural networks that take one kind of
thing and turn it into another: text into translated text, images into
captions. The pattern is always the same --- an \emph{encoder}
compresses the input into some intermediate representation, and a
\emph{decoder} unpacks that representation into the output.

In this notebook we're going to do something genuinely new. We're going
to build a network that can \textbf{generate} images: produce novel
faces it has never seen before. That's a different kind of problem than
translation or captioning. There's no input to translate from; we're
conjuring outputs from nothing.

The trick, it turns out, is to look at our familiar encoder/decoder
pattern from a strange angle. If we set things up just right, the
decoder alone can be used as a generator. The challenge is figuring out
what ``just right'' means. That's what the VAE is about.

Before we get there, let's review what we already know.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{§1. Three encoders we've already
built}\label{three-encoders-weve-already-built}

We've encountered encoders in three previous units. They look different
on the surface, but they're doing the same kind of thing.

\textbf{The RNN encoder (NMT).} We fed a source sentence one token at a
time and read off the final hidden state. That hidden state was a single
vector --- say, 512 numbers --- that ``summarized'' the entire sentence.
The decoder then unrolled this vector back into a target-language
sentence. Lossy compression: the source sentence has potentially
hundreds of words, but the hidden state has only 512 numbers.
Information had to be thrown away.

\textbf{The Transformer encoder (NMT, captioning).} Same idea but more
refined: instead of one vector for the whole input, we produced a vector
\emph{per input token}. So a 20-word sentence became 20 contextualized
vectors. Still lossy --- each output vector is much smaller than
``everything about that word in context'' --- but less aggressive than
the RNN's single-vector bottleneck.

\textbf{The ResNet encoder (captioning).} For images, we cut a
pretrained ResNet off before its classification head and used its
spatial feature maps as the encoding. A 224×224 image got compressed
into 49 feature vectors arranged in a 7×7 grid. Each vector summarizes a
region of the original image.

The common theme: \textbf{an encoder is lossy compression into a
representation a downstream task can use.} Different tasks like
different shapes of representation (one vector, many vectors, a grid),
but the underlying job is the same.

For the VAE, we're going to push compression even further. Each image
will be encoded to \emph{one} vector. For 64×64 face images, that's
12,288 pixel values being squeezed into 128 numbers --- a roughly 100×
compression. This is a much more aggressive bottleneck than the
captioning encoder used.

Why so aggressive? It'll be clear in §3 once we look at why we're doing
this in the first place.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{§2. Autoencoders}\label{autoencoders}

Suppose someone hands you the problem ``build a network that compresses
images and decompresses them back'' and asks you to design it from
scratch. What would you do?

Almost certainly this: build an encoder (compress to a small vector),
build a decoder (expand back to an image), train them together to
minimize the difference between the input and the reconstruction.

That's literally called an \textbf{autoencoder}, and it's been around
for decades. The encoder learns a useful compressed representation as a
side effect of the reconstruction task. No labels are needed --- the
image is its own supervision. (This is \textbf{unsupervised learning},
which is worth noting because most of what we've done this year has been
supervised.)

Autoencoders work fine for what they do. They give you compressed
representations that you can use for downstream tasks ---
classification, denoising, anomaly detection. The compressed vectors of
similar images tend to be near each other, which gives you a useful
``embedding'' of the data. None of this is generative, though. We've
built a compression tool, not a generator.

So how do we get from ``a thing that compresses and decompresses'' to
``a thing that generates new images?''

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{§3. Generators}\label{generators}

Here's the puzzle. We have a trained autoencoder. Decoder takes a
128-dimensional vector, produces a 64×64 face image. Encoder produces
such vectors from real images.

Could we just\ldots{} pick a random 128-dimensional vector and hand it
to the decoder?

\textbf{Try to predict what happens before reading on.}

What happens: garbage. The output is some noisy mess that doesn't look
like a face. Why?

Because the decoder was only ever trained on vectors that the encoder
produced. The encoder, in the course of training, mapped each real face
to some specific point in 128-dimensional space. Maybe all the faces
landed in some weird twisted region of that space --- a thin manifold,
or a few disconnected clusters, or who knows what shape. We don't know
what region, because we never asked the encoder to use any
\emph{particular} region. We just asked it to use \emph{some} region
that lets the decoder reconstruct.

When we pick a random 128-dim vector, we're almost certainly landing
somewhere the encoder never went, which means somewhere the decoder was
never trained. The decoder confidently produces nonsense for those
vectors.

Okay, what if we sampled from the \emph{actual} distribution of encoded
vectors? Encode the whole training set, look at where the encodings
landed, sample from there? In principle yes --- but practically, we'd
have to fit a high-dimensional density model to those encodings, which
is itself a hard generative modeling problem. We'd have just pushed the
problem back one level.

Here's the cleaner idea: \textbf{make the encoder produce vectors in
some distribution we already know how to sample from.} Pick a target
distribution that's easy to sample (say, N(0, 1) --- the standard
normal), and add a training pressure that pushes the encoder toward
producing outputs distributed that way.

If we can pull this off, generation becomes easy: sample z
\textasciitilde{} N(0, 1), feed it to the decoder, get an image.

This is the core idea of the VAE: don't just compress, \textbf{compress
into a sampleable space.} The ``variational'' part of the name refers to
a particular way of measuring how far the encoder's output distribution
is from our target.

The pieces we now need:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  A way to ask the encoder to produce outputs distributed like N(0, 1).
\item
  A way to measure when it's succeeded.
\item
  Confidence that this doesn't break reconstruction in the process.
\end{enumerate}

Let's tackle them in order.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{§4. Encoding to a
distribution}\label{encoding-to-a-distribution}

How do we ask the encoder to produce outputs distributed like
\(\mathcal{N}(0, 1)\)?

Here's a thought: the encoder is a deterministic function. It takes an
image, produces a vector. There's nothing random about it. If we run the
encoder on a batch of images, we get some specific set of vectors. We
could \emph{measure} how Gaussian-distributed those vectors look across
the batch, and add that to the loss.

This is actually a real approach --- it shows up in Adversarial
Autoencoders and Wasserstein Autoencoders. It works. But it requires
choosing some metric on distributions (adversarial training? Wasserstein
distance? matching moments?), and that metric has to be estimated from a
batch, which is statistically tricky in high dimensions.

The VAE takes a different approach. Instead of having the encoder output
a \emph{point} and then measuring the distribution of points across a
batch, \textbf{have the encoder output a \emph{distribution} directly.}

Here's how. For each input \(x\), the encoder produces two vectors: a
mean \texttt{mu} and a (log of the) variance \texttt{log\_var}. These
are interpreted as the parameters of a Gaussian distribution:

\[
q(z \mid x) = \mathcal{N}(\mu, \sigma^2), \qquad \sigma = \exp(0.5 \cdot \texttt{log\_var})
\]

Now we have a per-example distribution. Each \(x\) produces not a point
but a small Gaussian ``blob'' in latent space. When we want a latent
code for \(x\), we \emph{sample} from this Gaussian:

\[
z = \mu + \sigma \cdot \epsilon, \qquad \epsilon \sim \mathcal{N}(0, 1)
\]

And we feed \emph{that} sample to the decoder.

This setup might look ad-hoc. Let's pause and notice something subtle.

\textbf{The encoder doesn't ``describe'' a distribution. The training
procedure imposes one.} The \texttt{fc\_mu} and \texttt{fc\_log\_var}
layers in the encoder are just linear projections. There's nothing in
those layers that forces their outputs to be the parameters of any
particular distribution. We \emph{choose} to interpret them as Gaussian
parameters by the way we use them downstream. The sampling step
\(z = \mu + \sigma \cdot \epsilon\) is what makes the Gaussian
assumption real. Once \(z\) is generated this way, the model has no
choice but to be consistent with the Gaussian story, because the
gradients flow through that sampling formula.

The first time this hits, it can feel like a sleight of hand.
\emph{Wait, you just said these are means and variances, but they're
outputs of a \texttt{nn.Linear} --- what stops them from being garbage?}
The honest answer: nothing in the architecture stops it. The training
pressure does.

Why the sampling step? Two reasons that take a while to fully
appreciate.

\textbf{Reason 1: a distribution gives us something to push toward a
target.} If the encoder outputs a single point per \(x\), we can only
measure ``distribution-ness'' across a whole batch. If the encoder
outputs a \emph{distribution} per \(x\), we can measure the gap between
that per-example distribution and \(\mathcal{N}(0, 1)\) using a single,
closed-form formula (the KL divergence between two Gaussians, in §5).

\textbf{Reason 2: it acts as a regularizer.} Because \(z\) is sampled
each time we run a forward pass, the decoder sees a slightly different
\(z\) for the same \(x\) every time. Effectively, the decoder is being
asked: ``produce \(x\) correctly from \emph{any sample within a small
Gaussian neighborhood} of \(\mu\).'' That forces the decoder to be
smooth --- nearby latent points must decode to similar images, because
the decoder has been trained on a fuzzy region around each \(\mu\)
rather than a single point.

This second reason is genuinely surprising and worth sitting with. The
decoder is never trained on the deterministic encoding of \(x\). It's
trained on noisy samples around the deterministic encoding. The
randomness isn't a bug --- it's the entire mechanism by which the latent
space becomes smooth.

There's also the technical wrinkle of being able to compute gradients
through the sampling step. Naively,
\(z \sim \mathcal{N}(\mu, \sigma^2)\) isn't differentiable --- the
gradient can't flow back through the sampling operation to \(\mu\) and
\(\sigma\). The trick \(z = \mu + \sigma \cdot \epsilon\) (with
\(\epsilon\) drawn from a \emph{fixed} \(\mathcal{N}(0, 1)\) that we
don't differentiate through) puts \(\mu\) and \(\sigma\) into the
deterministic part of the computation. Now we can take
\(\partial \text{loss} / \partial \mu\) and
\(\partial \text{loss} / \partial \sigma\) like for any other
parameters. This is called the \textbf{reparameterization trick}, and
it's why the architecture above can be trained at all.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{§5. The KL term}\label{the-kl-term}

So we have an encoder that outputs (\texttt{mu}, \texttt{log\_var}) per
example. We sample \(z\) from \(\mathcal{N}(\mu, \sigma^2)\). The
decoder reconstructs from \(z\). So far the loss is just reconstruction
error.

Now we need to add a pressure that pushes \(q(z \mid x)\) toward
\(\mathcal{N}(0, 1)\). The natural measure of ``how far one distribution
is from another'' is the \textbf{Kullback-Leibler divergence}:

\[
\mathrm{KL}(q \| p) = \mathbb{E}_{z \sim q}[\log q(z) - \log p(z)]
\]

\(\mathrm{KL}\) is non-negative, equal to zero exactly when the two
distributions are identical. We want
\(\mathrm{KL}(q(z \mid x) \| \mathcal{N}(0, 1))\) to be small for every
\(x\).

For two Gaussians, \(\mathrm{KL}\) has a closed form. For
\(q(z \mid x) = \mathcal{N}(\mu, \sigma^2)\) and
\(p(z) = \mathcal{N}(0, 1)\):

\[
\mathrm{KL} = -\frac{1}{2} \sum \left(1 + \texttt{log\_var} - \mu^2 - \exp(\texttt{log\_var})\right)
\]

(summed over the latent dimensions). This is computable in a single line
of code. No sampling, no estimation --- closed-form formula in the
encoder's outputs.

So our total VAE loss is:

\[
L = \text{reconstruction loss} + \beta \cdot \mathrm{KL}\text{ term}
\]

where \(\beta\) is a hyperparameter. \(\beta = 1\) corresponds to the
``true'' theoretical objective (a lower bound on the log-likelihood of
the data, called the ELBO; we won't derive it but it's worth knowing it
exists). Smaller \(\beta\) puts more weight on reconstruction; larger
\(\beta\) puts more weight on matching the input.

That all sounds reasonable. But let's stop and look at it carefully,
because something seems off.

We're pushing every single \(q(z \mid x)\) to be close to
\(\mathcal{N}(0, 1)\). But if literally every per-example distribution
were \(\mathcal{N}(0, 1)\), then encoding any image would produce the
same distribution as encoding any other image. The decoder would receive
what amounts to fresh noise regardless of input. \emph{How could
reconstruction possibly work?}

Try to resolve this puzzle before reading on.

Here's the resolution. \textbf{In equilibrium, \(q(z \mid x)\) is
\emph{not} equal to \(\mathcal{N}(0, 1)\) for any \(x\).} The KL term
acts as a pressure pushing toward the input, but the reconstruction term
acts as an opposing pressure pulling away from it.

If \(\mu = 0\) and \(\texttt{log\_var} = 0\) (\(\sigma = 1\)) for every
input --- the values that make \(\mathrm{KL}\) exactly zero --- then
\(z\) is essentially noise unrelated to \(x\), and the decoder can't
possibly reconstruct anything. Reconstruction loss explodes. So the
optimizer can't sit there. It has to find a compromise.

The compromise looks like this: for each \(x\), \(\mu\) shifts away from
zero just enough to encode useful information about \(x\). \(\sigma\)
stays \emph{small} (well below 1) so that the blob is localized ---
nearby samples around \(\mu\) all roughly correspond to the same \(x\).
Each individual \(q(z \mid x)\) is a small Gaussian located
\emph{somewhere within} the \(\mathcal{N}(0, 1)\) region, not equal to
it.

The blobs from different x's land at different locations in latent
space. That's how the decoder distinguishes them.

\textbf{This is the picture worth carrying around:} Imagine
\(\mathcal{N}(0, 1)\) as a large bell-shaped region covering some
neighborhood of the origin in latent space. Each \(x\)'s blob is a
\emph{small marble} --- small in spread, located at a specific position
within the bell. Together, the marbles from the whole training set
scatter across the bell-shaped region and approximately fill it out.
Individually, no marble is the whole bell.

When we sample \(z \sim \mathcal{N}(0, 1)\) at generation time, we're
picking a random point in the bell. There's almost always \emph{some}
marble (or a region covered by overlapping marbles) near that point, and
the decoder has been trained to produce a plausible image there.

This picture also explains why the system works, but it's worth asking:
does the math \emph{guarantee} the marbles fill the bell? Or might they
cluster in one part and leave the rest empty?

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{§6. The aggregate-vs-per-example issue (optional deeper
dive)}\label{the-aggregate-vs-per-example-issue-optional-deeper-dive}

The KL term in our loss measures
\(\mathbb{E}_x[\mathrm{KL}(q(z \mid x) \| \mathcal{N}(0,1))]\) --- the
\emph{expected per-example} KL. But what we \emph{want} for generation
to work is for the \textbf{aggregate} distribution

\[
q(z) = \mathbb{E}_x[q(z \mid x)]
\]

(the mixture you get if you encode the whole dataset) to look like
\(\mathcal{N}(0, 1)\). These are not the same thing.

There's a nice identity that connects them. It can be shown that:

\[
\mathbb{E}_x[\mathrm{KL}(q(z \mid x) \| p(z))] = I(x; z) + \mathrm{KL}(q(z) \| p(z))
\]

where \(I(x; z)\) is the mutual information between input and latent
under the encoder's distribution. The proof is a one-line manipulation:
split \(\log(q(z \mid x)/p(z))\) into
\(\log(q(z \mid x)/q(z)) + \log(q(z)/p(z))\), take expectations, and the
first piece is mutual information while the second is aggregate
\(\mathrm{KL}\).

This identity is illuminating because it shows what the KL term is
\emph{really} doing:

\begin{itemize}
\tightlist
\item
  It's pushing \textbf{mutual information} down (toward zero information
  about x in z).
\item
  It's pushing \textbf{aggregate KL} down (toward the prior).
\end{itemize}

Both at once.

Reconstruction loss pushes back on mutual information --- the decoder
needs \(z\) to carry information about \(x\) to reconstruct. So in
equilibrium, mutual information stays positive (above whatever threshold
reconstruction requires), and aggregate \(\mathrm{KL}\) is non-zero (we
can't drive the whole KL term to zero without destroying mutual
information).

\textbf{The consequence:} the aggregate \(q(z)\) is approximately
\(\mathcal{N}(0, 1)\), but not exactly. There are ``holes'' in the
aggregate --- regions where the prior puts mass but few encoded points
actually land. This is called the \textbf{prior hole problem} or
\textbf{aggregate posterior gap}, and it's a well-known limitation of
VAEs.

You can see the consequences in samples: VAE-generated images sometimes
look ``off'' in subtle ways, and the worst samples come from regions of
the prior that are far from any cluster of training encodings. This is
part of why VAE samples have a characteristic blurriness/blandness
beyond what reconstruction MSE alone would predict.

A lot of post-VAE research (VampPrior, VQ-VAE, AAE, WAE, diffusion
models themselves) is essentially trying to address this gap. The VAE is
the simplest member of a family, not the final word.

For our purposes the takeaway is: \textbf{the convergence to
\(\mathcal{N}(0, 1)\) is approximate and bounded, not guaranteed and
tight.} The KL term in our loss is doing real work --- it's bounding the
aggregate \(\mathrm{KL}\) from above --- but it leaves a residual gap.
The intuition ``the loss nudges things toward \(\mathcal{N}(0, 1)\)'' is
precisely correct; what's bounded is \emph{how far} the aggregate can
wander.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{§7. KL annealing: a practical
wrinkle}\label{kl-annealing-a-practical-wrinkle}

There's a subtle failure mode worth knowing about. If we start training
with \(\beta = 1\) from the very first batch, the optimizer can find a
local minimum we don't want: every \(q(z \mid x)\) collapses exactly to
\(\mathcal{N}(0, 1)\), the decoder learns to produce a generic ``average
face'' regardless of \(z\), and reconstruction is bad but the KL term is
zero so the total loss is finite. Once the model is sitting in this
collapsed solution, it has trouble escaping.

This is called \textbf{posterior collapse}, and it's a real problem with
powerful decoders (especially text decoders that can fake plausible
outputs without conditioning on z).

The workaround: start with \(\beta = 0\) (pure reconstruction loss, the
model is just an autoencoder), and gradually ramp \(\beta\) up to 1 over
the first several epochs. The encoder is forced to use \(\mu\)
meaningfully because that's the only way to drive reconstruction down.
By the time KL pressure kicks in, the encoder is already in a regime
where ``use \(\mu\) informatively'' is the dominant solution, and
posterior collapse is no longer an easy escape.

This is called \textbf{KL annealing}, and you'll see it in the training
loop.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{§8. The decoder: getting from a vector back to an
image}\label{the-decoder-getting-from-a-vector-back-to-an-image}

We've spent a lot of time on the encoder side. The decoder is
conceptually simpler but architecturally interesting.

The shape problem: the encoder went from \texttt{64\ ×\ 64\ ×\ 3} (a
12,288-dim image) down to a 128-dim latent vector. The decoder has to go
in reverse --- from a 128-dim vector back to a 64 × 64 × 3 image. We
need to \emph{grow} spatial size.

How do we grow spatial size with a convolutional network?

A regular convolution with stride 2 \emph{shrinks} spatial size (halves
it). We need the opposite operation: something that \emph{doubles}
spatial size while doing a learned local operation. There's such a
thing, and it's called a \textbf{transposed convolution} (sometimes also
``deconvolution,'' though that name is misleading and the field has
mostly stopped using it).

Mechanically: a transposed convolution with stride 2 takes each input
cell and writes its kernel-sized footprint into a 2× larger output, with
overlapping regions summing. The result is an upsampling operation with
learnable weights --- the network learns \emph{how} to upsample, rather
than using a fixed rule like nearest-neighbor or bilinear interpolation.

Practical note: transposed convolutions can produce a characteristic
\textbf{checkerboard artifact} when the stride doesn't divide the kernel
size evenly. You may see faint checkerboard patterns in early-training
samples. Modern code often uses ``upsample then convolve'' instead
(nearest-neighbor upsample to 2× size, then a regular stride-1 conv) to
avoid this. For our notebook we'll use transposed convolutions because
they're simpler to read and the artifact is mild at 64 × 64.

The decoder architecture mirrors the encoder. The encoder went through
four stride-2 convolutions, shrinking spatial size from 64 to 4 (64 → 32
→ 16 → 8 → 4). The decoder does the reverse: starts with a linear
projection from the 128-dim latent to a 4 × 4 × 256 feature volume, then
four stride-2 transposed convolutions grow spatial size back up (4 → 8 →
16 → 32 → 64). The channel counts also reverse.

This mirror-image structure isn't required --- you could in principle
design any encoder/decoder pair --- but it's the cleanest, most
symmetric choice and works well.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{§9. Putting it all together}\label{putting-it-all-together}

Architecture:

\begin{itemize}
\tightlist
\item
  \textbf{Encoder:} four stride-2 convolutions on the image, then two
  parallel linear heads producing \texttt{mu} and \texttt{log\_var}.
\item
  \textbf{Reparameterization:}
  \(z = \mu + \exp(0.5 \cdot \texttt{log\_var}) \cdot \epsilon\), with
  \(\epsilon \sim \mathcal{N}(0, 1)\).
\item
  \textbf{Decoder:} linear projection to spatial feature volume, then
  four stride-2 transposed convolutions back up to image resolution,
  with Tanh final activation matching the {[}-1, 1{]} input
  normalization.
\end{itemize}

Loss:

\begin{itemize}
\tightlist
\item
  \textbf{Reconstruction:} MSE between original and reconstructed image,
  summed over pixels.
\item
  \textbf{KL:} closed-form expression in \texttt{mu} and
  \texttt{log\_var}, summed over latent dimensions.
\item
  \textbf{Total:} reconstruction + \(\beta \cdot \mathrm{KL}\), with
  \(\beta\) annealed from 0 to 1 over the first several epochs.
\end{itemize}

What to expect when you run the notebook:

\begin{itemize}
\tightlist
\item
  The loss drops fast at first while \(\beta\) is small (basically just
  an autoencoder phase).
\item
  When \(\beta\) ramps up, you'll see a visible inflection in the loss
  curve --- the model has to start balancing the two pressures.
\item
  Reconstructions get sharper over epochs but stay somewhat blurry. This
  blurriness isn't a bug you can fix by training longer; it's a
  fundamental consequence of MSE loss and the blob-not-point encoding
  (the decoder has to handle a neighborhood around each \(\mu\), and
  ends up producing the \emph{average} of plausible images for that
  neighborhood --- averaging plausible images produces blur).
\item
  Generated samples (sampling \(z \sim \mathcal{N}(0, 1)\) and decoding)
  will look like faces, but with the characteristic VAE-blurriness.
  They'll be visibly less sharp than the reconstructions.
\item
  Interpolations between two faces will produce smooth morphs through
  plausible intermediate faces. This is the payoff of the structured
  latent space --- interpolation in latent space corresponds to
  plausible-image interpolation in pixel space.
\item
  Varying a single latent dimension will sometimes reveal interpretable
  axes: hair color, lighting direction, pose, age. The model wasn't told
  about these attributes; they emerge from the structure of the data.
\end{itemize}

A few questions worth thinking about as you work through the notebook:

\textbf{What happens if you set \(\beta = 0\) for the entire run?} The
model becomes a regular autoencoder. Reconstructions should get sharper.
Generated samples (\(z \sim \mathcal{N}(0,1)\) and decode) should become
much worse --- possibly unrecognizable noise. Why? Because the encoder
is free to use any region of latent space, and \(\mathcal{N}(0, 1)\) is
unlikely to land where it actually put things.

\textbf{What would happen with \(\beta = 10\) instead of \(\beta = 1\)?}
More compression toward the prior. Reconstructions would get blurrier,
but generated samples might look \emph{more diverse} because the
aggregate is closer to a full \(\mathcal{N}(0, 1)\). This is the
trade-off space of ``\(\beta\)-VAEs.''

\textbf{The fundamental tension:} sharp reconstruction wants to use the
latent space freely; sampleable generation wants the latent space
compressed toward \(\mathcal{N}(0, 1)\). The \(\beta\) balances these.
There's no setting where both win without limit --- the trade-off is
real.

\begin{center}\rule{0.5\linewidth}{0.5pt}\end{center}

\subsection{§10. Looking ahead}\label{looking-ahead}

The VAE works. It generates faces. The faces are blurry, the latent
space isn't quite \(\mathcal{N}(0, 1)\), and there are theoretical loose
ends about the aggregate vs.~per-example distinction. None of these are
accidents --- they're consequences of the choices the VAE makes, and
each is a starting point for a research direction:

\begin{itemize}
\tightlist
\item
  \textbf{Sharper samples:} the blurriness comes from MSE-loss +
  averaging-over-blob. Diffusion models fix both by replacing the
  single-shot encode/decode with many small denoising steps. We'll see
  this next.
\item
  \textbf{Better latent fit:} VampPrior, VQ-VAE, and similar models
  replace the fixed \(\mathcal{N}(0, 1)\) prior with a learned one that
  better matches what the encoder actually produces.
\item
  \textbf{Direct aggregate matching:} Adversarial Autoencoders and
  Wasserstein Autoencoders skip the per-example KL and directly minimize
  a distance between the aggregate posterior and the prior. Sharper
  samples, harder training.
\end{itemize}

For now, sit with the VAE. The two ideas --- \emph{encoding to a
distribution} and \emph{KL pressure toward a sampleable prior} --- are
the conceptual core of most modern generative modeling, including
diffusion. If you understand the VAE in depth, the next several units
will feel like elaborations on themes you already know.

When you read through the notebook, try to keep the mental picture:

\begin{itemize}
\tightlist
\item
  A scatter of small Gaussian ``blobs'' tiling a region of latent space.
\item
  Each blob the encoding of one training image.
\item
  The decoder trained to produce that image from any sample inside the
  blob.
\item
  The blobs collectively shaped to approximately fill
  \(\mathcal{N}(0, 1)\).
\item
  Generation = drop a point into the bell-shaped region and decode.
\end{itemize}

If you can hold that picture and explain why each piece of the loss is
there, you've got the VAE.

\end{document}
